---
title: "California ARB Offset Credit Issuance Analysis"
author: "K. Reginald Wade"
date: "04/28/2022"
output: 
  slidy_presentation: 
    highlight: espresso
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Introduction

- ESG
- Fuel Regulation Standards
- California Cap and Trade
- Regulation and Risk Compliance

## Questions

- What projects should companies consider?
- What projects are at risk for invalidation?

## Required Packages
* tidyverse
    + readxl
    + dplyr
    + magrittr
    + ggplot2
* janitor

```{r message=FALSE, warning=FALSE}
#Install packages containing libraries: readxl, dplyr, magrittr, ggplot2
#install.packages("tidyverse")

#remove spaces and unique characters from column names
#install.packages("janitor")

library(tidyverse)
library(readxl)
library(dplyr, warn.conflicts = F)
library(magrittr)
library(ggplot2)
library(janitor)
```

## Source Data

The ARB Offset Credit Issuance data provides a list of offset projects that have been issued ARB offset credits.

```{r echo=TRUE}
#Read in data
arboc_issuance = read_excel("Data/arboc_issuance.xlsx", sheet="ARB Offset Credit Issuance", na=c("", "n/a"))
```

The issuance data is updated at 12PM PST on the 2nd and 4th business Wednesday of every month.

## Data Summary
```{r}
names(arboc_issuance)
```

## Data Cleaning
```{r echo=TRUE}
arboc_issuance_nice_names = clean_names(arboc_issuance)

names(arboc_issuance_nice_names)
```

```{r echo=TRUE}
#Create Project type description mapping
project_type = c("Forest", "ODS", "Livestock", "MMC", "Rice", "Urban")
project_type_desc = c("U.S. Forest Projects", "Ozone Depleting Substance Projects", "Livestock Projects", "Mine Methane Capture Projects", "Rice Cultivation Projects", "Urban Forest Projects")

project_types = data.frame(project_type, project_type_desc)
```

```{r echo=TRUE}
arboc_issuance_clean = inner_join(arboc_issuance_nice_names, project_types)

arboc_issuance_clean %>%
  top_n(5) %>% 
  select(c("carb_issuance_id", "project_type", "project_type", "project_type_desc", "offset_project_name", "start_of_invalidation_timeframe", "vintage"))
  
```
## Project Count Breakdown

```{r}
#arboc_issuance_clean %>%
#  group_by(project_type_desc) %>% 
#  count()
```

```{r}
arboc_issuance_clean %>%
  ggplot(aes(x=project_type, fill=project_type_desc)) + geom_bar()
```





## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

Some text about this table and how interesting the results are! 

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

some text about how cool my plot is. 

---

Three dashes make a new slide without having to put a title, and the three-# titles don't make a new slide, only the two and one. 

### see? 

still on the same slide. 
