---
title: "US Renewable Natural Gas Analysis"
author: "K. Reginald Wade"
date: "04/28/2022"
output: 
  slidy_presentation: 
    highlight: espresso
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Introduction

- ESG
- Fuel Regulation Standards
- Renewable Natural Gas (RNG)

## Questions

- What is the future outlook on RNG?
- What states are leading the charge?

## Required Packages
* tidyverse
    + readxl
    + dplyr
    + magrittr
    + ggplot2
* janitor
* data.table
* formattable
* plotly

```{r message=FALSE, warning=FALSE}
#Install packages containing libraries: readxl, dplyr, magrittr, ggplot2
#install.packages("tidyverse")

#remove spaces and unique characters from column names
#install.packages("janitor")

#Nice data tables
#install.packages("data.table")
#install.packages("formattable")

library(tidyverse)
library(readxl)
library(dplyr)
library(magrittr)
library(ggplot2)
library(janitor)
library(data.table)
library(formattable)
library(plotly)
```

## Source Data

### LFG Energy Project Data Files (March 2022)

```{r echo=TRUE}
#Read in data
landfillgas_raw = read_excel("Data/lmopdata.xlsx", sheet="LMOP Database")
```

Source: https://www.epa.gov/lmop/lmop-landfill-and-project-database

### AgSTAR Livestock Anaerobic Digester Database

```{r echo=TRUE}
anaerobicdigesterfas_raw = read_excel("Data/agstar-livestock-ad-database.xlsx", sheet="Operational and Construction")
```

AgSTAR publishes updated data on this website periodically (generally, two times per year) to enhance public access to information and support the development of biogas recovery projects

Source: https://www.epa.gov/agstar/livestock-anaerobic-digester-database

### RNG Facility Data

```{r echo=TRUE}
rng_facility_raw = read_csv("Data/RNG_Facilities.csv", show_col_types = FALSE)
```


## Data Summary

### LMOP Landfill and Landfill Gas Energy Project Database

```{r echo=TRUE}
names(landfillgas_raw)
```

### AgSTAR Anaerobic Digester Database

```{r echo=TRUE}
names(anaerobicdigesterfas_raw)
```

```{r echo=TRUE}
names(rng_facility_raw)
```

## Data Cleaning

### LMOP Landfill and Landfill Gas Energy Project Database

```{r echo=TRUE}
landfillgas_clean = clean_names(landfillgas_raw)
names(landfillgas_clean)
```

### AgSTAR Anaerobic Digester Database

```{r echo=TRUE}
anaerobicdigesterfas_clean = clean_names(anaerobicdigesterfas_raw)
names(anaerobicdigesterfas_clean)
```

### RNG Facility Data

```{r echo=TRUE}
rng_facility_clean = clean_names(rng_facility_raw)
names(rng_facility_clean)
```

## EDA

### LMOP Landfill and Landfill Gas Energy Project Database

```{r}
landfillgas_clean %>% 
  select(landfill_name, project_name, state, waste_in_place_tons, lfg_energy_project_type, project_start_date) %>% 
  head(5) %>% 
  formattable()
```

### AgSTAR Anaerobic Digester Database

```{r}
anaerobicdigesterfas_clean %>%
  select(project_name, state, project_type, animal_farm_type_s, year_operational, biogas_end_use_s) %>% 
  head(5) %>% 
  formattable()
```

## Project Breakdown

### RNG Opearational Project Count by Year

```{r}
landfillgas_count_by_year = landfillgas_clean %>% 
  filter(current_project_status == "Operational") %>% 
  mutate(rng_type = "Landfill RNG", operational_year = format(as.Date(project_start_date, format="%d/%m/%Y"),"%Y")) %>% 
  select(rng_type, operational_year) %>% 
  group_by(rng_type, operational_year) %>% 
  count()

anaerobicdigesterfas_count_by_year = anaerobicdigesterfas_clean %>%
  mutate(rng_type = "Ag RNG", operational_year = format(as.Date(ISOdate(year_operational, 1, 1), format="%d/%m/%Y"),"%Y")) %>% 
  select(rng_type, operational_year) %>%
  group_by(rng_type, operational_year) %>%
  count()

rng_count_by_year = bind_rows(landfillgas_count_by_year, anaerobicdigesterfas_count_by_year)

rng_count_by_year %>%
  filter(operational_year > 2002) %>% 
  ggplot(aes(x=as.character(operational_year), y=n, fill=rng_type)) + geom_bar(stat='identity') + labs(title="Operational Project Count by Year", fill="RNG Type", x="Operatonal Year", y="Project Count") + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
```

### RNG Operational Project Count by State

```{r}
landfillgas_count_by_state = landfillgas_clean %>% 
  filter(current_project_status == "Operational") %>% 
  mutate(rng_type = "Landfill RNG") %>% 
  select(rng_type, state) %>% 
  group_by(rng_type, state) %>% 
  count()

anaerobicdigesterfas_count_by_state = anaerobicdigesterfas_clean %>%
  mutate(rng_type = "Ag RNG") %>% 
  select(rng_type, state) %>%
  group_by(rng_type, state) %>%
  count()

rng_count_by_state = bind_rows(landfillgas_count_by_state, anaerobicdigesterfas_count_by_state)

rng_count_by_state %>%
  ggplot(aes(x=reorder(state, n), y=n, fill=rng_type, )) + geom_bar(stat='identity') + labs(title="Operational Project Count by State", fill="RNG Type", x="State", y="Project Count") + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5)) + coord_flip()
```

```{r}
landfillgas_count_by_state = landfillgas_clean %>% 
  filter(current_project_status == "Operational") %>% 
  mutate(rng_type = "Landfill RNG") %>% 
  select(rng_type, state) %>% 
  group_by(rng_type, state) %>% 
  count()

anaerobicdigesterfas_count_by_state = anaerobicdigesterfas_clean %>%
  mutate(rng_type = "Ag RNG") %>% 
  select(rng_type, state) %>%
  group_by(rng_type, state) %>%
  count()

rng_count_by_state = bind_rows(landfillgas_count_by_state, anaerobicdigesterfas_count_by_state)

rng_count_by_state %>%
  ggplot(aes(x=reorder(state, n), y=n, fill=rng_type, )) + geom_bar(stat='identity') + labs(title="Project Count by State", fill="RNG Type", x="State", y="Project Count") + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5)) + coord_flip() + facet_wrap(~rng_type)
```

## Landfill RNG Project Map

```{r}
project_graph_lf = plot_geo(landfillgas_count_by_state,
                         locationmode="USA-states") %>%
  add_trace(locations = ~state,
            z = ~n,
            zmin = 0, 
            zmax = max(rng_count_by_state$n),
            color= ~n,
            colorscale = "Viridis",
            reversescale = T) %>% 
  layout(title = "Landfill RNG Project Map by State", geo = list(scope = "usa"))

project_graph_lf
```

## Ag Gas Project Map

```{r}
project_graph_ag = plot_geo(anaerobicdigesterfas_count_by_state,
                         locationmode="USA-states") %>%
  add_trace(locations = ~state,
            z = ~n,
            zmin = 0, 
            zmax = max(rng_count_by_state$n),
            color= ~n,
            colorscale = "Viridis",
            reversescale = T) %>% 
  layout(title = "Ag RNG Project Map by State", geo = list(scope = "usa"))

project_graph_ag
```
```{r}
landfillgas_count_by_geo = landfillgas_clean %>% 
  filter(current_project_status == "Operational") %>% 
  select(landfill_name, state, latitude, longitude) %>% 
  group_by(landfill_name, state, latitude, longitude) %>% 
  count()
```

## RNG Facility Locations

```{r warning=FALSE}
g = list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("gray85"),
  subunitwidth = 1,
  countrywidth = 1,
  subunitcolor = toRGB("white"),
  countrycolor = toRGB("white")
)

project_graph_geo = plot_geo(rng_facility_clean,
               locationmode = "USA-states",
               sizes = c(1, 250)) %>% 
  add_markers(x = ~longitude, y = ~latitude,
              size = ~ifelse(is.na(gas_flow_to_project_scfm), 4000, gas_flow_to_project_scfm), color = ~city,
              hoverinfo = "text", fill = ~"",
              text = ~paste(rng_facility_clean$facility_name, " - ", rng_facility_clean$city)) %>%
  layout(title = "RNG Facilities", geo = g, showlegend = FALSE)
  

project_graph_geo
```
## Landfill RNG Project Forecast

```{r message=FALSE, warning=FALSE}
options(scipen=10000)

landfillgas_clean %>% 
  filter(!current_project_status  %in% c("Shutdown", "Unknown", "Low Potential")) %>% 
  filter(project_type_category  != "Unknown") %>% 
  group_by(current_project_status, project_type_category) %>% 
  summarise(mean_waste =mean(waste_in_place_tons, na.rm=T), n = sum()) %>% 
  ggplot(aes(x=reorder(current_project_status, mean_waste), y=mean_waste, fill=project_type_category)) + geom_bar(stat="identity", position="dodge") + labs(title="Average Waste by Project Status", fill="Project Type Category", y="Average Waste (tons)", x="Project Status")
```
```{r eval=FALSE, include=FALSE}
landfillgas_facilities = landfillgas_clean %>% 
  filter(current_project_status == "Operational", current_landfill_status =="Open") %>% 
  mutate(operational_year = as.numeric(format(as.Date(project_start_date, format="%d/%m/%Y"),"%Y"))) %>%
  select(operational_year) %>% 
  group_by(operational_year) %>% 
  count() %>% 
  mutate(id=1, facilities = n) %>% 
  select(id, operational_year, facilities)

get_num_facilities = function(year) {
  
  filter(landfillgas_facilities, operational_year <= year) %>%
    group_by(id) %>% 
    summarize(num_facilities = sum(facilities)) %>% 
    select(num_facilities) %>% 
    as.numeric()
}

landfillgas_facilities_by_year = landfillgas_facilities %>% 
  mutate(rng_type = "Landfill RNG", num_facilities = get_num_facilities(operational_year)) %>% 
  select(rng_type, operational_year, num_facilities)

anaerobicdigesterfas_facilities_by_year = anaerobicdigesterfas_clean %>%
  mutate(rng_type = "Ag RNG", operational_year = as.numeric(format(as.Date(ISOdate(year_operational, 1, 1), format="%d/%m/%Y"),"%Y"))) %>% 
  select(rng_type, operational_year) %>%
  group_by(rng_type, operational_year) %>%
  count() %>% 
  mutate(num_facilities = n) %>% 
  select(rng_type, operational_year, num_facilities)

rng_facilities_by_year = bind_rows(landfillgas_facilities_by_year, anaerobicdigesterfas_facilities_by_year)

rng_facilities_by_year %>%
  #filter(operational_year > 2002) %>% 
  ggplot(aes(x=operational_year, y=num_facilities, fill=rng_type)) + geom_bar(stat='identity') + labs(title="Operational Project Count by Year", fill="RNG Type", x="Operatonal Year", y="Project Count") + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
```

## Ag RNG Production Forecast

```{r warning=FALSE}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational") %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=animal_farm_type_s, y=num_animals, fill=project_type)) + geom_bar(position="dodge", stat = "identity") + scale_y_log10()
```

### Bio Gas Generation Estimate/day (cubic ft)

```{r warning=FALSE}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational") %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=year_operational, y=biogas_generation_estimate_cu_ft_day, color=animal_farm_type_s)) + geom_point() + scale_x_log10() + scale_y_log10()
```

### Electricity Generation Estimate/year (KWH)

```{r warning=FALSE}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational") %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=year_operational, y=electricity_generated_k_wh_yr, color=animal_farm_type_s)) + geom_jitter() + scale_x_log10() + scale_y_log10()
```

## Carbon Zero

### Ag Emissions Reductions

```{r warning=FALSE}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational") %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=year_operational, y=total_emission_reductions_mtco2e_yr, color=animal_farm_type_s)) + geom_jitter() + scale_x_log10() + scale_y_log10()
```

### Landfill Gas Emissions Reductions

```{r}
landfillgas_clean %>% 
  filter(current_landfill_status == "Open", current_project_status == "Operational") %>% 
  select(project_start_date, waste_in_place_tons, lfg_collected_mmscfd, lfg_energy_project_type, project_type_category, current_year_emission_reductions_mmtco2e_yr_direct, actual_mw_generation, current_year_emission_reductions_mmtco2e_yr_avoided) %>% 
  ggplot(aes(x=project_start_date, y=current_year_emission_reductions_mmtco2e_yr_direct, color=project_type_category)) + geom_jitter()
```

### Landfill Gas Emissions Avoidance

```{r}
landfillgas_clean %>% 
  filter(current_landfill_status == "Open", current_project_status == "Operational") %>% 
  select(project_start_date, waste_in_place_tons, lfg_collected_mmscfd, lfg_energy_project_type, project_type_category, current_year_emission_reductions_mmtco2e_yr_direct, actual_mw_generation, current_year_emission_reductions_mmtco2e_yr_avoided) %>% 
  ggplot(aes(x=project_start_date, y=current_year_emission_reductions_mmtco2e_yr_avoided, color=project_type_category)) + geom_jitter()
```


