---
title: "US Renewable Natural Gas Analysis Report"
author: "Kweku Reginald Wade"
date: "04/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Introduction

In the last three years there has been a seismic shift in what investors value. No longer are they looking to corporations for merely a return on investment, but a return on social capital. Social Capital has been defined in many ways by many people, but I define social capital in terms how corporations treat the Environment, Society, and how they Govern their employees.Taking the first initial from the words: Environment, Society, and Governance, yields the acronym ESG. ESG has become the de-facto measure investors use to grade corporations' social capital.

In terms of the E in ESG, investors are looking to invest in renewable projects. Once such project type is Renewable Natural Gas. Renewable Natural Gas is natural gas that comes from non-fossil sources, such as landfills and farms. Landfills can products methane, the key component in natural gas, that is released by means of decaying bio-material found in trash. Farm animals also release methane through bodily waste that can be captured in an anaerobic digestion system.

My report seeks to answer four questions for investors:

- What is the future outlook on RNG?
- What states are leading the charge?
- What is the demand for RNG?
- Does it make a difference on the environment?

# Required Packages

The following libraries were essential in order to complete my project:

* tidyverse     - Used for reading, formatting, and plotting data
    + readxl
    + dplyr
    + magrittr
    + ggplot2
* janitor       - Used for cleaning data
* data.table    - Used for formatting data tables
* formattable   - Used for formatting data tables
* plotly        - Used for plotting data

```{r message=FALSE, warning=FALSE}
#Loading essential libraries
library(tidyverse)
library(readxl)
library(dplyr)
library(magrittr)
library(ggplot2)
library(janitor)
library(data.table)
library(formattable)
library(plotly)
```

# Source Data

The source of my data originated from the Environmental Protection Agency, better known by it's acronym, the EPA. The EPA provides a wealth of data on renewable natural gas projects in both the Landfill and the farm animal flavor. The farm animal projects are better known as AgStar projects.

The landfill data will be loaded from the LFG Energy Project Data Files as of March 2022. The location of this file can be located from the following URL: Source: https://www.epa.gov/lmop/lmop-landfill-and-project-database

```{r echo=TRUE}
#Loading in landfill project data
landfillgas_raw = read_excel("Data/lmopdata.xlsx", sheet="LMOP Database")
```

The AgSTAR Livestock Anaerobic Digester Database can be loaded from the following URL: https://www.epa.gov/agstar/livestock-anaerobic-digester-database

AgSTAR publishes updated data on this website periodically (generally, two times per year) to enhance public access to information and support the development of biogas recovery projects

```{r echo=TRUE}
#Loading AgStar project data
anaerobicdigesterfas_raw = read_excel("Data/agstar-livestock-ad-database.xlsx", sheet="Operational and Construction")
```

The Renewable Natural Gas, or RNG, facility data contains details of 173 facilities around the United States.

```{r echo=TRUE}
#Loading RNG Facility data
rng_facility_raw = read_csv("Data/RNG_Facilities.csv", show_col_types = FALSE)
```

# Data Summary

The data describes the physical and geo location of landfill in the United States. The data also contains project information such as when the project started, the amount of wastes in tons the landfill holds, the amount of gas and, or amount of electricity that the landfill produces as well as emission reduction and avoidance data. The Landfill data consists of the following schema:

```{r echo=TRUE}
#Show names of the columns
names(landfillgas_raw)
```

The data describes the physical location of Anaerobic digesters in the United States. The data also contains project information such as when the farm become operational, the type of animal/farm, the amount of gas and, or amount of electricity that the farm produces as well as emission reduction data. The AgSTAR Anaerobic Digester data consists of the following schema:

```{r echo=TRUE}
#Show names of the columns
names(anaerobicdigesterfas_raw)
```

The data describes the physical and geo-location of RNG facilities in the United States. The data also contains the amount of gas the facility produces.

The RNG facility data consists of the following schema:

```{r echo=TRUE}
#Show names of the columns
names(rng_facility_raw)
```

# Data Cleaning

As you have seen from the previous schema, many of the column names consist of spaces. We now use the janitor library to transform the columns names so that any special character will be replaces by an underscore.

The following removes special characters from the Landfill columns and replaces them with underscores.

```{r echo=TRUE}
#Clean landfill data columns
landfillgas_clean = clean_names(landfillgas_raw)
names(landfillgas_clean)
```

The following removes special characters from the AgSTAR Anaerobic Digester Data columns and replaces them with underscores.

```{r echo=TRUE}
#Clean AgStart Anaerobic Digester data columns
anaerobicdigesterfas_clean = clean_names(anaerobicdigesterfas_raw)
names(anaerobicdigesterfas_clean)
```

The following removes special characters from the RNG facility data columns and replaces them with underscores.

```{r echo=TRUE}
#Clean RNG Facility data columns
rng_facility_clean = clean_names(rng_facility_raw)
names(rng_facility_clean)
```

# EDA

In our first step at exploratory data analysis we look at the first 5 rows of the Landfill. A few columns of note is the project name of the landfill, its location, the amount of waste, the project type and the start date of the project.

```{r}
landfillgas_clean %>% 
  select(landfill_name, project_name, state, waste_in_place_tons, lfg_energy_project_type, project_start_date) %>% 
  head(5) %>% 
  formattable()
```

We also look at the first 5 rows of the AgSTAR Anaerobic Digester data. A few columns of note is the project name of the digester, its location, the project type, the type of animal on the farm, the year of operation, and the end use of the bio gas.

```{r}
anaerobicdigesterfas_clean %>%
  select(project_name, state, project_type, animal_farm_type_s, year_operational, biogas_end_use_s) %>% 
  head(5) %>% 
  formattable()
```

# Project Breakdown

We now look at RNG project breakdown by operational year. Here we see how projects have ramped up over the years. We also see the increase in AgStar projects year by year. 

```{r}
landfillgas_count_by_year = landfillgas_clean %>% 
  filter(current_project_status == "Operational") %>% 
  mutate(rng_type = "Landfill RNG", operational_year = format(as.Date(project_start_date, format="%d/%m/%Y"),"%Y")) %>% 
  select(rng_type, operational_year) %>% 
  group_by(rng_type, operational_year) %>% 
  count()

anaerobicdigesterfas_count_by_year = anaerobicdigesterfas_clean %>%
  mutate(rng_type = "Ag RNG", operational_year = format(as.Date(ISOdate(year_operational, 1, 1), format="%d/%m/%Y"),"%Y")) %>% 
  select(rng_type, operational_year) %>%
  group_by(rng_type, operational_year) %>%
  count()

rng_count_by_year = bind_rows(landfillgas_count_by_year, anaerobicdigesterfas_count_by_year)

rng_count_by_year %>%
  filter(operational_year > 2002) %>% 
  ggplot(aes(x=as.character(operational_year), y=n, fill=rng_type)) + geom_bar(stat='identity') + labs(title="Operational Project Count by Year", fill="RNG Type", x="Operatonal Year", y="Project Count") + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
```

In looking at RNG the operational project count by state, we see California leading the charge, followed by Pennsylvania and New York.

```{r}
landfillgas_count_by_state = landfillgas_clean %>% 
  filter(current_project_status == "Operational") %>% 
  mutate(rng_type = "Landfill RNG") %>% 
  select(rng_type, state) %>% 
  group_by(rng_type, state) %>% 
  count()

anaerobicdigesterfas_count_by_state = anaerobicdigesterfas_clean %>%
  mutate(rng_type = "Ag RNG") %>% 
  select(rng_type, state) %>%
  group_by(rng_type, state) %>%
  count()

rng_count_by_state = bind_rows(landfillgas_count_by_state, anaerobicdigesterfas_count_by_state)

rng_count_by_state %>%
  ggplot(aes(x=reorder(state, n), y=n, fill=rng_type, )) + geom_bar(stat='identity') + labs(title="Operational Project Count by State", fill="RNG Type", x="State", y="Project Count") + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5)) + coord_flip()
```

Project Type Breakdown

```{r}
landfillgas_count_by_state = landfillgas_clean %>% 
  filter(current_project_status == "Operational") %>% 
  mutate(rng_type = "Landfill RNG") %>% 
  select(rng_type, state) %>% 
  group_by(rng_type, state) %>% 
  count()

anaerobicdigesterfas_count_by_state = anaerobicdigesterfas_clean %>%
  mutate(rng_type = "Ag RNG") %>% 
  select(rng_type, state) %>%
  group_by(rng_type, state) %>%
  count()

rng_count_by_state = bind_rows(landfillgas_count_by_state, anaerobicdigesterfas_count_by_state)

rng_count_by_state %>%
  ggplot(aes(x=reorder(state, n), y=n, fill=rng_type, )) + geom_bar(stat='identity') + labs(title="Project Count by State", fill="RNG Type", x="State", y="Project Count") + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5)) + coord_flip() + facet_wrap(~rng_type)
```

From a map perspective, we see a heat map of the number of landfill projects by state. Again California leads the charge in terms of landfills, followed by Michigan. We also notice Wyoming does not have any landfills that are allocated for landfill biogas projects.

```{r}
project_graph_lf = plot_geo(landfillgas_count_by_state,
                         locationmode="USA-states") %>%
  add_trace(locations = ~state,
            z = ~n,
            zmin = 0, 
            zmax = max(rng_count_by_state$n),
            color= ~n,
            colorscale = "Viridis",
            reversescale = T) %>% 
  layout(title = "Landfill RNG Project Map by State", geo = list(scope = "usa"))

project_graph_lf
```

From a map perspective, we see a heat map of the number of digester projects by state. Again California leads the charge in terms of digesters, followed by Wisconsin and New York. We also notice the absence of digester projects for states such as Nevada, New Mexico, and surprisingly Louisiana.

```{r}
project_graph_ag = plot_geo(anaerobicdigesterfas_count_by_state,
                         locationmode="USA-states") %>%
  add_trace(locations = ~state,
            z = ~n,
            zmin = 0, 
            zmax = max(rng_count_by_state$n),
            color= ~n,
            colorscale = "Viridis",
            reversescale = T) %>% 
  layout(title = "Ag RNG Project Map by State", geo = list(scope = "usa"))

project_graph_ag
```
```{r}
landfillgas_count_by_geo = landfillgas_clean %>% 
  filter(current_project_status == "Operational") %>% 
  select(landfill_name, state, latitude, longitude) %>% 
  group_by(landfill_name, state, latitude, longitude) %>% 
  count()
```

## RNG Facility Locations

```{r warning=FALSE}
g = list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("gray85"),
  subunitwidth = 1,
  countrywidth = 1,
  subunitcolor = toRGB("white"),
  countrycolor = toRGB("white")
)

project_graph_geo = plot_geo(rng_facility_clean,
               locationmode = "USA-states",
               sizes = c(1, 250)) %>% 
  add_markers(x = ~longitude, y = ~latitude,
              size = ~ifelse(is.na(gas_flow_to_project_scfm), 4000, gas_flow_to_project_scfm), color = ~city,
              hoverinfo = "text", fill = ~"",
              text = ~paste(rng_facility_clean$facility_name, " - ", rng_facility_clean$city)) %>%
  layout(title = "RNG Facilities", geo = g, showlegend = FALSE)
  

project_graph_geo
```

# Landfill RNG Project Forecast

The outlook for Landfill RNG projects looks bright. We see over 30 Billion tons of waste candidate projects, as well as over 40 Billion tons of waste landfills that are either in the planning state or in currently being constructed.

```{r message=FALSE, warning=FALSE}
options(scipen=10000)

landfillgas_clean %>% 
  filter(!current_project_status  %in% c("Shutdown", "Unknown", "Low Potential")) %>% 
  filter(project_type_category  != "Unknown") %>% 
  group_by(current_project_status, project_type_category) %>% 
  summarise(mean_waste =mean(waste_in_place_tons, na.rm=T), n = sum()) %>% 
  ggplot(aes(x=reorder(current_project_status, mean_waste), y=mean_waste, fill=project_type_category)) + geom_bar(stat="identity", position="dodge") + labs(title="Average Waste by Project Status", fill="Project Type Category", y="Average Waste (tons)", x="Project Status")
```
```{r eval=FALSE, include=FALSE}
landfillgas_facilities = landfillgas_clean %>% 
  filter(current_project_status == "Operational", current_landfill_status =="Open") %>% 
  mutate(operational_year = as.numeric(format(as.Date(project_start_date, format="%d/%m/%Y"),"%Y"))) %>%
  select(operational_year) %>% 
  group_by(operational_year) %>% 
  count() %>% 
  mutate(id=1, facilities = n) %>% 
  select(id, operational_year, facilities)

get_num_facilities = function(year) {
  
  filter(landfillgas_facilities, operational_year <= year) %>%
    group_by(id) %>% 
    summarize(num_facilities = sum(facilities)) %>% 
    select(num_facilities) %>% 
    as.numeric()
}

landfillgas_facilities_by_year = landfillgas_facilities %>% 
  mutate(rng_type = "Landfill RNG", num_facilities = get_num_facilities(operational_year)) %>% 
  select(rng_type, operational_year, num_facilities)

anaerobicdigesterfas_facilities_by_year = anaerobicdigesterfas_clean %>%
  mutate(rng_type = "Ag RNG", operational_year = as.numeric(format(as.Date(ISOdate(year_operational, 1, 1), format="%d/%m/%Y"),"%Y"))) %>% 
  select(rng_type, operational_year) %>%
  group_by(rng_type, operational_year) %>%
  count() %>% 
  mutate(num_facilities = n) %>% 
  select(rng_type, operational_year, num_facilities)

rng_facilities_by_year = bind_rows(landfillgas_facilities_by_year, anaerobicdigesterfas_facilities_by_year)

rng_facilities_by_year %>%
  #filter(operational_year > 2002) %>% 
  ggplot(aes(x=operational_year, y=num_facilities, fill=rng_type)) + geom_bar(stat='identity') + labs(title="Operational Project Count by Year", fill="RNG Type", x="Operatonal Year", y="Project Count") + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
```

# Ag RNG Production Forecast

The outlook for AgStar projects looks bright as well. We see over 100,000 animals being used to produce biogas.

```{r warning=FALSE}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational") %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=animal_farm_type_s, y=num_animals, fill=project_type)) + geom_bar(position="dodge", stat = "identity") + scale_y_log10() + labs(title="Operational Project Count by Year", fill="Project Type", x="Animal Type", y="Animals")
```

# Bio Gas Generation Estimate/day (cubic ft)

```{r warning=FALSE}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational") %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=year_operational, y=biogas_generation_estimate_cu_ft_day, color=animal_farm_type_s)) + geom_point() + scale_x_log10() + scale_y_log10() + labs(title="Bio Gas Generation Estimate/day (cubic ft)", color="Animal Farm Type", x="Operational Year", y="Bio Gas Generation Estimate")
```

# Electricity Generation Estimate/year (KWH)

```{r warning=FALSE}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational") %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=year_operational, y=electricity_generated_k_wh_yr, color=animal_farm_type_s)) + geom_jitter() + scale_x_log10() + scale_y_log10() + labs(title="Electricity Generation Estimate/year (KWH)", color="Animal Farm Type", x="Operational Year", y="Electricity Generation Estimate")
```

# Demand

## Project Count by End Use

```{r}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  filter(biogas_end_use_s %in% c("CNG", "Electricity", "Cogeneration", "Pipeline Gas", "Pipeline to Electricity")) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>%
  ggplot(aes(biogas_end_use_s, fill=animal_farm_type_s)) + geom_bar(position="dodge") + labs(title="Project Count by End Use", x="Biogas End Use", y="Projects", fill="Animal Type")
```

## Power Generation

```{r}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational", year_operational > 2010, receiving_utility != "NA") %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  filter(!is.na(electricity_generated_k_wh_yr)) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=reorder(substr(receiving_utility, 1, 30), electricity_generated_k_wh_yr), y=electricity_generated_k_wh_yr, fill=animal_farm_type_s)) + geom_bar(stat = "identity") + labs(title="Utility Company Generation Acquisition", fill="Animal Farm Type", x="Utility", y="Electricity Generated") + coord_flip()
```

# Carbon Zero

## Ag Emissions Reductions

```{r warning=FALSE}
anaerobicdigesterfas_clean %>% 
  mutate(num_animals = pmax(dairy, cattle, poultry, swine, na.rm=T)) %>% 
  filter(status == "Operational") %>% 
  filter(animal_farm_type_s %in% c("Dairy", "Cattle", "Poultry", "Swine")) %>% 
  select(project_name, status, state, animal_farm_type_s, num_animals, project_type, digester_type, year_operational, biogas_generation_estimate_cu_ft_day, electricity_generated_k_wh_yr, biogas_end_use_s, receiving_utility, total_emission_reductions_mtco2e_yr) %>% 
  ggplot(aes(x=year_operational, y=total_emission_reductions_mtco2e_yr, color=animal_farm_type_s)) + geom_jitter() + scale_x_log10() + scale_y_log10() + labs(title="Ag Emissions Reductions", color="Animal Farm Type", x="Operational Year", y="Total Emission Reductions")
```

## Landfill Gas Emissions Reductions

```{r}
landfillgas_clean %>% 
  filter(current_landfill_status == "Open", current_project_status == "Operational") %>% 
  select(project_start_date, waste_in_place_tons, lfg_collected_mmscfd, lfg_energy_project_type, project_type_category, current_year_emission_reductions_mmtco2e_yr_direct, actual_mw_generation, current_year_emission_reductions_mmtco2e_yr_avoided) %>% 
  ggplot(aes(x=project_start_date, y=current_year_emission_reductions_mmtco2e_yr_direct * 1000, color=project_type_category)) + geom_jitter() + labs(title="Landfill Gas Emissions Reductions", color="Product Type Category", x="Operational Year", y="Total Emission")
```

## Landfill Gas Emissions Avoidance

```{r}
landfillgas_clean %>% 
  filter(current_landfill_status == "Open", current_project_status == "Operational") %>% 
  select(project_start_date, waste_in_place_tons, lfg_collected_mmscfd, lfg_energy_project_type, project_type_category, current_year_emission_reductions_mmtco2e_yr_direct, actual_mw_generation, current_year_emission_reductions_mmtco2e_yr_avoided) %>% 
  ggplot(aes(x=project_start_date, y=current_year_emission_reductions_mmtco2e_yr_avoided * 1000, color=project_type_category)) + geom_jitter() + labs(title="Landfill Gas Emissions Avoidance", color="Product Type Category", x="Operational Year", y="Total Emission")
```

# Summary

In looking at the data I believe the outlook for RNG is positive. Through the direction of the EPA we see states such as California, Michigan and New York leading the charge. We even see conservative states like Texas investing in RNG projects. 

Demand is ramping up as municipalities such as Dallas leading the way in buying compressed natural gas vehicles. We also see every major electricity utility adding renewable natural gas to their portfolio in-lieu of fossil fuel based natural gas.

As we embark on the road to Carbon zero, we see the use of RNG has an effect as data shows an increase in not only emission reduction, but avoidance.
