---
title: "US Renewable Natural Gas Analysis"
author: "K. Reginald Wade"
date: "04/28/2022"
output: 
  slidy_presentation: 
    highlight: espresso
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Introduction

- ESG
- Fuel Regulation Standards
- Renewable Natural Gas (RNG)

## Questions

- What is the future outlook on RNG?
- What states are leading the charge?

## Required Packages
* tidyverse
    + readxl
    + dplyr
    + magrittr
    + ggplot2
* janitor
* data.table
* formattable

```{r message=FALSE, warning=FALSE}
#Install packages containing libraries: readxl, dplyr, magrittr, ggplot2
#install.packages("tidyverse")

#remove spaces and unique characters from column names
#install.packages("janitor")

#Nice data tables
#install.packages("data.table")
#install.packages("formattable")

library(tidyverse)
library(readxl)
library(dplyr)
library(magrittr)
library(ggplot2)
library(janitor)
library(data.table)
library(formattable)
```

## Source Data

### LFG Energy Project Data Files (March 2022)

```{r echo=TRUE}
#Read in data
landfillgas_raw = read_excel("Data/opprjslmopdata-detailed.xlsx", sheet="LMOP Database")
```

Source: https://www.epa.gov/lmop/lmop-landfill-and-project-database

### AgSTAR Livestock Anaerobic Digester Database

```{r echo=TRUE}
anaerobicdigesterfas_raw = read_excel("Data/agstar-livestock-ad-database.xlsx", sheet="Operational and Construction")
```

AgSTAR publishes updated data on this website periodically (generally, two times per year) to enhance public access to information and support the development of biogas recovery projects

Source: https://www.epa.gov/agstar/livestock-anaerobic-digester-database

## Data Summary

### LMOP Landfill and Landfill Gas Energy Project Database

```{r echo=TRUE}
names(landfillgas_raw)
```

### AgSTAR Anaerobic Digester Database

```{r echo=TRUE}
names(anaerobicdigesterfas_raw)
```

## Data Cleaning

### LMOP Landfill and Landfill Gas Energy Project Database

```{r echo=TRUE}
landfillgas_clean = clean_names(landfillgas_raw)
names(landfillgas_clean)
```

### AgSTAR Anaerobic Digester Database

```{r echo=TRUE}
anaerobicdigesterfas_clean = clean_names(anaerobicdigesterfas_raw)
names(anaerobicdigesterfas_clean)
```

## EDA

### LMOP Landfill and Landfill Gas Energy Project Database

```{r}
landfillgas_clean %>% 
  select(landfill_name) %>% 
  head(5) %>% 
  formattable()
```

### AgSTAR Anaerobic Digester Database

```{r}
anaerobicdigesterfas_clean %>%
  select(project_name, project_type, animal_farm_type_s, year_operational, biogas_end_use_s) %>% 
  head(5) %>% 
  formattable()
```


### Project Breakdown
```{r}

```


